### 2022-09-03 main/chapter_searches.tex
rearrange some quotes, add some likelihood fisher content

Modern frequentist methods are largely attributable to work by
Ronald A. Fisher (for likelihood, $p$-values, and estimation criteria~\cite{
fisher1921probable,
fisher1922estimators,
fisher1925smrw,
fisher1956statistical
}),
and Jerzy Neyman and Egon S. Pearson
(for statistical hypothesis testing and confidence intervals~\cite{
neymanpearson1933lemma,
neymanpearson1928max
})
in the 1920s and 30s.
Together, the Fisher and Neyman-Pearson methods do not form a harmonious
framework, but comprise a loose collection of ideas and practices that survive
in modern science, and explain important features of the methods we apply to
LHC data, which Section~\ref{sec:searches_practice} develops further.

Some literature names this Fisher/Neyman-Pearson assemblage
``classical'' statistics~\cite{lehmann2011fisher, Feldman:1997qc}, but that is
simply inaccurate for a 20th century divergence from 18th century
ideas~\cite{bayes1763lii, laplace1774stigler}.
Jaynes instead names them ``orthodox'~\cite{jaynes2003probability},
which may be accurate in some contexts, but alternative
(heretical?) ideas have taken hold under modern LHC authorities ---
the orthodoxy has moved on.
We therefore stick to the name `frequentist', as it is a firm common factor.

\begin{figure}[tp]
\centering
\includegraphics[width=0.95\textwidth]{figures/seacrhes_fisher_likelihood_binomial.png}
\caption[
Ratio thresholds of a Binomial likelihood function, reproduced from Fisher
]{%
Ratio thresholds of a Binomial likelihood function, reproduced from
Fisher~\cite{fisher1956statistical}.
Levels are drawn at $1/2$, $1/5$, and $1/15$ of its maximum.
Their extents could be reported in a table as a practical communication of
the likelihood function's profile.
}
\label{fig:searches_fisher_likelihood}
\end{figure}

% likelihood
Fisher coined the term `likelihood'~\cite{fisher1921probable}, in the usage
we have employed, for the probability of observed data given varied
contexts\footnote{%
Fisher used the term `likelihood' for the function exploring
contexts and `probability' for the distribution over
outcomes~\cite{fisher1925smrw}.
We choose the alternative word `prior' for his `probability' since the
likelihood is also a probability, just not a probability \emph{distribution}.
},
and deserves credit for popularising its use in interpreting and communicating
data.
% figure here TODO
% likelihood quotes
Fisher forcefully argued against the use of Bayes' rule and prior
probabilities, stating for example that
\begin{singlespacing}
\begin{displayquote}
We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations~\cite{
fisher1921probable
}
\end{displayquote}
\end{singlespacing}

% p-values

@@

\TODO{Fisher's great contribution --- likelihood}
\TODO{
Fisher:
criteria for estimators,
(consistency, efficiency, sufficiency)
~\cite{fisher1922estimators}
testing, p-value,
likelihood, also ratios of p-values (here named probable errors)
(introduced the term likelihood? cite?)
~\cite{fisher1921probable}
tail area = significance
~\cite{fisher1925smrw}
}
Figure~\ref{fig:searches_fisher_likelihood}

\TODO{
Fisher:
``We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations''
~\cite{fisher1921probable}
Fisher:
``absurdly academic''
~\cite{fisher1956statistical}
% \begin{epigraphs}
% \qitem{%
% We can know nothing of the probability of hypotheses or hypothetical
% quantities.
% On the other hand we may ascertain the likelihood of hypotheses and
% hypothetical quantities by calculation from observations%
% }%
% {Ronald A. Fisher, 1921~\cite{fisher1921probable}}
% \end{epigraphs}
There is something horrifying in the ideological movement represented by the
doctrine that reasoning, properly speaking, cannot be applied to empirical data
to lead to inferences valid in the real world.
~\cite{fisher1956statistical}
}

\begin{singlespacing}
\begin{displayquote}
We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations
\end{displayquote}
\end{singlespacing}
~\cite{fisher1921probable}

\begin{singlespacing}
\begin{displayquote}
For all purposes, and more particularly for the \textit{communication} of the
relevant evidence supplied by a body of data, the values of the Mathematical
Likelihood are better fitted to analyse, summarize, and communicate
statistical evidence of types too weak to supply true probability statements
\end{displayquote}
\end{singlespacing}
~\cite{fisher1956statistical}

@@

\TODO{
Nayman-Pearson:
Seek good test statistics --
likelihood ratio intuitive for two-hypothesis comparison
prove it optimal in a reasonable sense
idea of long-run mostly not wrong
recede to maximum likelihood ratios
~\cite{neymanpearson1928max}
``for which hypothesis A is more probable and those for which it is less
probable (sic!)''
unbiased (unbiassed) test differs from unbiasedness of estimator
``Then the two corresponding densities at the sample point
are (Al) 6-42 x 10-20, (A2) 5 18*x 1O-4 and the ratio of the likelihood of Hypothesis
Al to- that of A2 is 1P24 x 10-16. This ratio confirms the common-sense judgment
that A2 is far more plausible than A1''
\^ Another posterior argument --- this likelihood ratio in in fact a change
in plausibilities. For example, if A2 is an inconsistent, self-contradictory
theory, it is totally implausible, and not more plausible despite predicting
these data better.
``The statistician can
balance the numerical verdict of likelihood against the vaguer expectations de-
rived from `a priori' considerations, and it must be left to his judgment to
at what point the evidence in favour of alternative hypotheses becomes so con-
vincing that Hypothesis A must be rejected.''
Introduce maximum likelihood ratio as interesting without claiming it to be
necessarily best.
Neyman-Pearson lemma --- reproduce Fig 5?
~\cite{neymanpearson1933lemma}
}

>>>>

Modern frequentist methods are largely attributable to work by
Ronald A. Fisher (for likelihood, $p$-values, and estimation criteria~\cite{
fisher1921probable,
fisher1922estimators,
fisher1925smrw,
fisher1956statistical
}),
and Jerzy Neyman and Egon S. Pearson
(for statistical hypothesis testing and confidence intervals~\cite{
neymanpearson1933lemma,
neymanpearson1928max
})
in the 1920s and 30s.
These Fisher and Neyman-Pearson schools do not form a unified framework, but
do comprise a collection of ideas and practices that survive in modern science,
and which explain some methods, to be introduced in
Section~\ref{sec:searches_practice}, that we apply to modern LHC data.

Some literature names this Fisher/Neyman-Pearson assemblage
``classical'' statistics~\cite{lehmann2011fisher, Feldman:1997qc}, but that is
simply inaccurate for a 20th century innovation created to replace 18th century
ideas~\cite{bayes1763lii, laplace1774stigler}.
Jaynes instead names them ``orthodox''~\cite{jaynes2003probability},
which may be accurate in some contexts, but alternative
(heretical?) ideas have taken hold in research at the LHC ---
the orthodoxy has moved on.
We therefore stick with `frequentist'.

\begin{figure}[tp]
\centering
\includegraphics[width=0.95\textwidth]{figures/seacrhes_fisher_likelihood_binomial.png}
\caption[
Ratio thresholds of a Binomial likelihood function, reproduced from Fisher
]{%
Ratio thresholds of a Binomial likelihood function, reproduced from
Fisher~\cite{fisher1956statistical}.
Horizontal lines mark regions where the likelihood exceeds $1/2$, $1/5$, or
$1/15$ of its maximum respectively.
The extents of these lines could be tabulated to usefully communicate
aspects of its profile.
}
\label{fig:searches_fisher_likelihood}
\end{figure}

\paragraph{Fisher.}
% likelihood
Fisher coined the term `likelihood'~\cite{fisher1921probable} in the usage that
we employ for the probability of observed data given varied
contexts%
\footnote{%
Fisher used the term `likelihood' for the function exploring
contexts and `probability' for the distribution over
outcomes~\cite{fisher1925smrw}.
We choose the alternative word `prior' for his `probability' since the
likelihood is also a probability, just not a probability \emph{distribution}.
},
and deserves credit for popularising its use in interpreting and communicating
data without the confounding influence of prior probabilities.
In his own words,%
\begin{displayquote}
\small
``%
We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations%
''~\cite{fisher1921probable},
\end{displayquote}
and furthermore we agree that in comparison to confidence intervals,
\begin{displayquote}
``%
For all purposes, and more particularly for the \textit{communication} of the
relevant evidence supplied by a body of data, the values of the Mathematical
Likelihood are better fitted to analyse, summarize, and communicate
statistical evidence of types too weak to supply true probability statements%
''~\cite{fisher1956statistical}.
\end{displayquote}
For example, Figure~\ref{fig:searches_fisher_likelihood} displays a figure by
Fisher~\cite{fisher1956statistical} that illustrates how data can be usefully
reported by their likelihood's maximum and the regions in which it exceeds
certain ratios relative to that maximum.%
\footnote{%
We use such $L(\check{x})\pm\mathrm{ratio}$ visualizations for
$\pm1\sigma$ and $\pm2\sigma$ error bars on Poisson data in
Figure~7 of~\cite{lester2022hunting}.
When making error bar recommendations, \atlas' statistics committee appears
to have not considered this option~\cite{cowan2011errorbars};
I cannot explain why not, since our statistics experts must surely be familiar
with Fisher's work.
}

% likelihood quotes
In avoiding the pitfalls of posteriors, Fisher's use of likelihoods remains
innovative and practical.
However, the hard rejection of priors on hypotheses also prevents calculation
of evidence values, without which likelihoods can become confounded
in too many parameters (perhaps making them nuisances).
To circumvent this, frequentist methods employ two key concepts:
test statistics, which can reduce the data to remove unwanted parameters,
and $p$-values, which divert the discourse
from what the alternative hypotheses \emph{are}
to what alternative data \emph{might have been}.

% test statistics
A test statistic is a function of data.
All real-valued functions are permitted, but some are more useful than others.
\TODO{TODO t-statistic}

% p-values
(probable error?), known in modern jargon as $p$-values

@@

\TODO{Fisher's great contribution --- likelihood}
\TODO{
Fisher:
criteria for estimators,
(consistency, efficiency, sufficiency)
~\cite{fisher1922estimators}
testing, p-value,
likelihood, also ratios of p-values (here named probable errors)
(introduced the term likelihood? cite?)
~\cite{fisher1921probable}
tail area = significance
~\cite{fisher1925smrw}
}

\TODO{
Fisher:
``absurdly academic''
~\cite{fisher1956statistical}
There is something horrifying in the ideological movement represented by the
doctrine that reasoning, properly speaking, cannot be applied to empirical data
to lead to inferences valid in the real world.
~\cite{fisher1956statistical}
}

@@

\paragraph{Neyman-Pearson.}
\TODO{
Nayman-Pearson:
Seek good test statistics --
likelihood ratio intuitive for two-hypothesis comparison
prove it optimal in a reasonable sense
idea of long-run mostly not wrong
recede to maximum likelihood ratios
~\cite{neymanpearson1928max}
``for which hypothesis A is more probable and those for which it is less
probable (sic!)''
unbiased (unbiassed) test differs from unbiasedness of estimator
``Then the two corresponding densities at the sample point
are (Al) 6-42 x 10-20, (A2) 5 18*x 1O-4 and the ratio of the likelihood of Hypothesis
Al to- that of A2 is 1P24 x 10-16. This ratio confirms the common-sense judgment
that A2 is far more plausible than A1''
\^ Another posterior argument --- this likelihood ratio in in fact a change
in plausibilities. For example, if A2 is an inconsistent, self-contradictory
theory, it is totally implausible, and not more plausible despite predicting
these data better.
``The statistician can
balance the numerical verdict of likelihood against the vaguer expectations de-
rived from `a priori' considerations, and it must be left to his judgment to
at what point the evidence in favour of alternative hypotheses becomes so con-
vincing that Hypothesis A must be rejected.''
Introduce maximum likelihood ratio as interesting without claiming it to be
necessarily best.
Neyman-Pearson lemma --- reproduce Fig 5?
~\cite{neymanpearson1933lemma}
}
